ODE and Diffusion Models of Tumor Growth

### 2024-03-15 paper.tex
11

\begin{equation}
    c_{total}(t) = c_0 + c_{prod}(t) - n(t) - n_{lost}(t)
\end{equation}
where

@@

>>>>

\begin{equation}
    c_{total}(t) = c_0 + c_{prod}(t) - n(t) - n_{lost}(t)
    \label{eq:conservation}
\end{equation}
where

@@

where $3\lambda$ is the proportionality constant.

Our final equation requires more rigorous mathematics. We know that the proliferation rate is parameterized
by both $\sigma$ and $\beta$. Since we need to accumulate the total amount of cells produced over a continuous
time period, we must integrate. However, there are some values of $t$ where $r_1(t) > r_2(t)$, so we must include
another integral to account for this. Mathematically, this is:

\begin{equation}
    c_{prod}(t) = 4\pi \int_{0}^{t} dt \int_{max(r_1(t), r_2(t))}^{r_3(t)} S(\sigma, \beta) r^2 dr
\end{equation}

where the factor of 1/3 seen in the other equations comes from the integration of the $r^2$ term.

We must now plug these formulas into the conservation of mass equation ($\ref{eq:conservation}$) seen above.

$ r_3^3(t) - r_1^3(t) = r_3^3(0) + 3\int_{0}^{t} dt \int_{max(r_1(t), r_2(t))}^{r_3(t)} S(\sigma, \beta) r^2 dr -
    r_1^3(t) - \int_{0}^{t}3\lambda r_1^3(t) dt $ (removing the factors of $4\pi$)

\begin{equation}
    r_3^3(t) = r_3^3(0) + 3\int_{0}^{t} dt \int_{max(r_1(t), r_2(t))}^{r_3(t)} S(\sigma, \beta) r^2 dr
    - \int_{0}^{t}3\lambda r_1^3(t) dt
\end{equation}
This is equation (3.1) in Greenspan's paper. It is described as the `conservation of volume' equation and is what
governs the growth of the tumor. To see how the tumor changes with time, the time partial derivative of this equation
would be useful. This is equation (3.2) in the paper and will be verified below:

Recall that taking the derivative of an integral with a constant lower bound and $t$ as the upper bound removes the integral,
simplifying the second and third terms on the right hand side of the equation. So we get:
\begin{equation}
    3r_3^2\frac{dr_3}{dt} = 3\int_{max(r_1(t), r_2(t))}^{r_3(t)} S(\sigma, \beta) r^2 dr - 3\lambda r_1^3(t)
\end{equation}

Removing the factor of 3 from each term simplifies the equation to (3.2).

\subsection{The Inward Movement of Live Tumor Cells}

Using the same concepts, it is also possible to model the movement of cells towards the inside of the necrotic
center. This rate is equal to the rate of production minus the volume rate of death of live cells at the center.
If $v$ is the velocity of a cell towards the center, the relative velocity of this cell with respect to the necrotic core
can be written as $r_1'(t) - v$. See the following figure for clarification:

\begin{figure}[ht]
    \centering
    \begin{tikzpicture}[scale=2]
        \draw (0, 0) circle (1);
        \draw (0, 0) circle (0.6);
        \draw (0, 0) circle (0.4);

        \draw[->, thick] (0, 0.4) -- (0, 0.6) node[left, above] {$r_1'(t)$};
        \draw[->, thick] (1, 0) -- (0, 0) node[midway, above] {$-v$};
    \end{tikzpicture}
    \caption{A cross-section of the sphere that represents the inner layer of the tumor. If the tumor is growing outward (the positive direction) and the velocity of the tumor is towards the inside,
        then the magnitude of the relative velocity of the cell volume is smaller than the rate $v$.}
    \label{fig:necrosis}
\end{figure}

The rate of change of the outer volume of cells moving into the necrotic core will then be $-4\pi r_1^2(r_1'(t) - v)$.

So the rate of change of the total cell volume, $c_{total}(t)$ is then

\begin{equation}
    \frac{4\pi}{3} \frac{d}{dt}(r_3^3(t)-r_1^3(t)) = \int_{max(r_1, r_2)}^{r_3} S(\sigma, \beta)r^2 dt -4\pi r_1^2(r_1'(t) - v)
\end{equation}

To find the value of $v$, we need to compare this equation with equation (25). Taking the derivative of the left hand side of (26) yields:

$4\pi(r_3^2 \frac{dr_3}{dt} - r_1^2 \frac{dr_1}{dt})$.

Moving the term $-4\pi r_1^2 \frac{dr_1}{dt}$ to the right-hand side cancels it out. The first term on the left and right hand sides
of both equations will be the same. So the only term not identical to equation (25) can be set equal to the respective one in equation (26):

\begin{equation}
    4\pi r_1^2v = -\lambda r_1^3
\end{equation}

with solution

\begin{equation}
    v = -\lambda r_1
\end{equation}

\subsection{Overview of Diffusion Model}

Greenspan (1972) proposed that the steady state of the tumor was caused by the combination of two factors: the lack of nutrients and the forces caused by necrosis
(as described above.) In that model, the inhibitor $\beta$ was assumed to be a product of necrosis. Following that, we need to make a few approximations
before describing the model:

\begin{enumerate}
    \item[(i)] $\beta(r, t)$ is produced in the core ($r < r_1(t)$) at some constant rate per unit volume $P$.
    \item[(ii)] The rates of diffusion of any of the chemicals are uniformly constant throughout all three regions of the tumor. As seen in Hill (1928), we will use $k$ and $k'$ to denote the diffusion constants for $\sigma$ and $\beta$ respectively.
    \item[(iii)] As also seen in Hill's model, cells consume nutrients at a constant rate per volume $a$.
    \item[(iv)] The rate of growth of cells per unit volume is some constant $s$. Greenspan (1972) uses an approximation for the distribution of new cells via mitosis: \begin{equation}
              S(\sigma, \beta) = sH(\sigma-\sigma_{min})H(\beta_{max}-\beta)
          \end{equation}
          where $H(x) = 1, x \geq 0; H(x) = 0, x < 0$.
    \item[(v)] The composition of the tumor medium is kept constant and the compositions outside the tumor (at the outer surface) are constants as well.
\end{enumerate}

Using the nice simplification given by assumption (iv), we can simplify our previous equation further:

\begin{equation}
    r_3^2\frac{dr_3}{dt} = \int_{max(r_1(t), r_2(t))}^{r_3(t)} s r^2 dr - \lambda r_1^3(t)
\end{equation}

Evaluating the integral:

\begin{equation}
    r_3^2\frac{dr_3}{dt} = \frac{s}{3} (r_3^3(t) - max(r_1^3(t), r_2^3(t))) - \lambda r_1^3(t)
\end{equation}

This is our diffusion equation.
